\bf An algebraic formulation of dependent type theory

### 2014-11-07 Presentation/presentation.tex
comparison with B-systems added

\newcommand\subgoal[1]{\begin{huge}\begin{center}\textbf{Subgoal: {\color{teal!90!black}#1}}\end{center}\end{huge}}
\newcommand\important[1]{\textbf{\color{red!90!black}#1}}

@@

\begin{frame}
\frametitle{\bf Goal and acknowledgements}
In this presentation, an essentially algebraic theory of type dependency will be given. Thus, we will
do two things:
\begin{itemize}
\item Give a syntactic description of the theory in the usual style of
name-free type theory.
\item Define the corresponding algebraic objects, called \important{E-systems}, in a category with finite limits.
\end{itemize}
\pause
For many insights in the current presentation I am grateful to
\begin{itemize}
\item Vladimir Voevodsky
\item Richard Garner
\item Steve Awodey
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{\bf Consequences of being algebraic}
The requirement that the theory is essentially algebraic gives immediate access to the standard tools of algebra, such as
\begin{itemize}
\item free algebras;
\item quotients algebras.
\end{itemize}
\pause
The major source of examples of E-systems are C-systems.
\\[\baselineskip]
\pause
In a way similar to the definition of E-systems, it is possible to define
\important{internal E-systems}. One of the aims of this research was to have
a systematic treatment of internal models.
\begin{itemize}
\item Nothing more than plain dependent type theory is needed to express what
an internal E-system is.
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{\bf The basic judgments}
\begin{align*}
\jalign\jctx{\Gamma} 
& \jalign\jctxeq{\Gamma}{\Gamma'}
  \\
\jalign\jfam{\Gamma}{A} 
& \jalign\jfameq{\Gamma}{A}{B}
  \\
\jalign\jterm{\Gamma}{A}{a} 
& \jalign\jtermeq{\Gamma}{A}{a}{y}.
\end{align*}
\\[\baselineskip]
There are rules expressing that all three kinds of judgmental equality are
convertible equivalence relations.

@@

\begin{comment}
\begin{frame}
\frametitle{\bf Rules: judgmental equality is an equivalence relation}
\begin{footnotesize}
\begin{align*}
& \inference
  { \jctx{\Gamma}
    }
  { \jctxeq{\Gamma}{\Gamma}
    } 
& & \inference
    { \jctxeq{\Gamma}{\Delta}
      }
    { \jctxeq{\Delta}{\Gamma}
      } 
& & \inference
    { \jctxeq{\Gamma}{\Delta}
      \jctxeq{\Delta}{\greek{E}}
      }
    { \jctxeq{\Gamma}{\greek{E}}
      }
    \\
& \inference
  { \jfam{\Gamma}{A}
    }
  { \jfameq{\Gamma}{A}{A}
    } 
& & \inference
    { \jfameq{\Gamma}{A}{B}
      }
    { \jfameq{\Gamma}{B}{A}
      }
& & \inference
    { \jfameq{\Gamma}{A}{B}
      \jfameq{\Gamma}{B}{C}
      }
    { \jfameq{\Gamma}{A}{C}
      }
    \\
& \inference
  { \jterm{\Gamma}{A}{a}
    }
  { \jtermeq{\Gamma}{A}{a}{a}
    }
& & \inference
    { \jtermeq{\Gamma}{A}{a}{y}
      }
    { \jtermeq{\Gamma}{A}{y}{a}
      }
& & \inference
    { \jtermeq{\Gamma}{A}{a}{y}
      \jtermeq{\Gamma}{A}{y}{z}
      }
    { \jtermeq{\Gamma}{A}{a}{z}
      }
\end{align*}
\begin{align*}
& \inference
  { \jctxeq{\Gamma}{\Delta}
    \jfam{\Gamma}{A}
    }
  { \jfam{\Delta}{A}
    }
& & \inference
    { \jctxeq{\Gamma}{\Delta}
      \jfameq{\Gamma}{A}{B}
      }
    { \jfameq{\Delta}{A}{B}
      }
    \\
& \inference
  { \jctxeq{\Gamma}{\Delta}
    \jterm{\Gamma}{A}{a}
    }
  { \jterm{\Delta}{A}{a}
    }
& & \inference
    { \jctxeq{\Gamma}{\Delta}
      \jtermeq{\Gamma}{A}{a}{y}
      }
    { \jtermeq{\Delta}{A}{a}{y}
      }
    \\
& \inference
  { \jfameq{\Gamma}{A}{B}
    \jterm{\Gamma}{A}{a}
    }
  { \jterm{\Gamma}{B}{a}
    }
& & \inference
    { \jfameq{\Gamma}{A}{B}
      \jtermeq{\Gamma}{A}{a}{y}
      }
    { \jtermeq{\Gamma}{B}{a}{y}
      }
\end{align*}
\end{footnotesize}
\end{frame}
\end{comment}

\begin{frame}
\frametitle{\bf The fundamental structure of E-systems}
The \important{fundamental structure of an E-system $\stesys$} in a category with finite limits consists of
\begin{equation*}
\begin{tikzcd}
\stesyst
  \ar{d}{\ebd}
  \\
\stesysf
  \ar{d}{\eft}
  \\
\stesysc
\end{tikzcd}
\end{equation*}
\pause
\begin{itemize}
\item The object $\stesysc$ represents the sort of contexts.
\pause
\item The object $\stesysf$ represents the sort of families.
\pause
\item Every family has a context, this is represented by $\eft:\stesysf\to\stesysc$.
\pause
\item The object $\stesyst$ represents the sort of terms.
\pause
\item Every term has a family associated to it, this is represented by
$\partial:\stesyst\to\stesysf$.
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{\bf Mixed commutative diagrams}
Suppose $f:\stesys\to\stesys'$ is a pre-extension homomorphism. We say that
a diagram
\begin{equation*}
\begin{tikzcd}[ampersand replacement=\&]
\stesys
  \ar{r}{f}
  \ar{d}[swap]{p}
  \&
\stesys'
  \ar{d}{p'}
  \\
X \ar{r}[swap]{g}
  \&
Y
\end{tikzcd}
\end{equation*}
commutes if the diagram
\begin{equation*}
\begin{tikzcd}[ampersand replacement=\&]
\stesysc
  \ar{r}{f_0}
  \ar{d}[swap]{p}
  \&
\stesysc'
  \ar{d}{p'}
  \\
X \ar{r}[swap]{g}
  \&
Y
\end{tikzcd}
\end{equation*}
commutes.
\end{frame}

@@

\begin{frame}
\frametitle{\bf Rules for the theory of projections}
Introduction rules for units:
\begin{align*}
& \inference
  { \jfam{\Gamma}{A}
    }
  { \unfoldall{\jhom{\Gamma}{A}{A}{\idtm{A}}}
    }
& & \inference
    { \jfameq{\Gamma}{A}{A'}
      }
    { \unfoldall{\jhomeq{\Gamma}{A}{A}{\idtm{A}}{\idtm{A'}}}
      }
\end{align*}
Weakening preserves the unit:
\begin{equation*}
\inference
  { \jfam{\Gamma}{A}
    \jfam{{\Gamma}{B}}{Q}
    }
  { \unfoldall{\jhomeq
      {{{\Gamma}{A}}{\ctxwk{A}{B}}}
      {\ctxwk{A}{Q}}
      {\ctxwk{A}{Q}}
      {\ctxwk{A}{\idtm{Q}}}
      {\idtm{\ctxwk{A}{Q}}}
      }
    }
\end{equation*}
\end{frame}

>>>>

\newcommand\subgoal[1]{\begin{huge}\begin{center}\textbf{Subgoal: {\color{teal!90!black}#1}}\end{center}\end{huge}}
\newcommand\important[1]{\textbf{\color{red!90!black}#1}}
\newcommand{\Btilde}{\skew{4}{\tilde}{B}}

@@

\begin{frame}
\frametitle{\bf Goal and acknowledgements}
We will do two things:
\begin{itemize}
\item Give a syntactic reformulation of type theory in the usual style of
name-free type theory.
\item Define the corresponding algebraic objects, called \important{E-systems}, in a category with finite limits.
\end{itemize}
\pause
For many insights in the current presentation I am grateful to
\begin{itemize}
\item Vladimir Voevodsky
\item Richard Garner
\item Steve Awodey
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{\bf The scope of the theory}
The requirement that the theory is essentially algebraic gives immediate access to the standard tools of algebra, such as
\begin{itemize}
\item free algebras;
\item quotients algebras.
\end{itemize}
\pause
The major source of examples of E-systems are \important{B-systems}.
\begin{itemize}
\item B-systems are stratified E-systems.
\item There is a full and faithful functor from C-systems to pre-B-systems with the B-systems as its image.
\end{itemize}
\pause
In a way similar to the definition of E-systems, it is possible to define
\important{internal E-systems}. One of the aims of this research was to have
a systematic treatment of internal models.
\begin{itemize}
\item Nothing more than plain dependent type theory is needed to express what
an internal E-system is.
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{\bf The basic judgments}
\begin{align*}
\jalign\jctx{\Gamma} 
& \jalign\jctxeq{\Gamma}{\Delta}
  \\
\jalign\jfam{\Gamma}{A} 
& \jalign\jfameq{\Gamma}{A}{B}
  \\
\jalign\jterm{\Gamma}{A}{a} 
& \jalign\jtermeq{\Gamma}{A}{a}{b}.
\end{align*}
\\[\baselineskip]
There are rules expressing that all three kinds of judgmental equality are
convertible equivalence relations.

@@

\begin{frame}
\frametitle{\bf The fundamental structure of E-systems}
The \important{fundamental structure of an E-system $\stesys$} in a category with finite limits consists of
\begin{equation*}
\begin{tikzcd}
\stesyst
  \ar{d}{\ebd}
  \\
\stesysf
  \ar{d}{\eft}
  \\
\stesysc
\end{tikzcd}
\end{equation*}
\pause
\begin{itemize}
\item The object $\stesysc$ represents the sort of contexts.
\pause
\item The object $\stesysf$ represents the sort of families.
\pause
\item Every family has a context, this is represented by $\eft:\stesysf\to\stesysc$.
\pause
\item The object $\stesyst$ represents the sort of terms.
\pause
\item Every term has a family associated to it, this is represented by
$\ebd:\stesyst\to\stesysf$.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{\bf Comparison with B-systems}
\begin{minipage}[t]{.42\textwidth}
The fundamental structure of E-systems:
\\[1.8cm]
\begin{equation*}
\begin{tikzcd}[ampersand replacement=\&]
\stesyst
  \ar{dr}[swap]{\ebd}
  \\
  {}\&
\stesysf
  \ar{d}{\eft}
  \\
  {}\&
\stesysc
\end{tikzcd}
\end{equation*}
\end{minipage}
\hfill
\begin{minipage}[t]{.42\textwidth}
The underlying structure of B-systems:
\vfill
\begin{equation*}
\begin{tikzcd}[ampersand replacement=\&]
\Btilde_2
  \ar{dr}[swap]{\ebd}
  \&
\vdots
  \ar{d}{\eft}
  \\
\Btilde_1
  \ar{dr}[swap]{\ebd}
  \&
B_2
  \ar{d}{\eft}
  \\
\Btilde_0
  \ar{dr}[swap]{\ebd}
  \&
B_1
  \ar{d}{\eft}
  \\
  {}\&
B_0
\end{tikzcd}
\end{equation*}
\end{minipage}
\end{frame}

@@

\begin{frame}
\frametitle{\bf Condensed commutative diagrams}
Suppose $f:\stesys\to\stesys'$ is a pre-extension homomorphism. We say that
a diagram
\begin{equation*}
\begin{tikzcd}[ampersand replacement=\&]
\stesys
  \ar{r}{f}
  \ar{d}[swap]{p}
  \&
\stesys'
  \ar{d}{p'}
  \\
X \ar{r}[swap]{g}
  \&
Y
\end{tikzcd}
\end{equation*}
commutes if the diagram
\begin{equation*}
\begin{tikzcd}[ampersand replacement=\&]
\stesysc
  \ar{r}{f_0}
  \ar{d}[swap]{p}
  \&
\stesysc'
  \ar{d}{p'}
  \\
X \ar{r}[swap]{g}
  \&
Y
\end{tikzcd}
\end{equation*}
commutes.
\end{frame}

@@

\begin{frame}
\frametitle{\bf Rules for the theory of projections}
Introduction rules for units:
\begin{align*}
& \inference
  { \jfam{\Gamma}{A}
    }
  { \unfoldall{\jhom{\Gamma}{A}{A}{\idtm{A}}}
    }
& & \inference
    { \jfameq{\Gamma}{A}{A'}
      }
    { \unfoldall{\jhomeq{\Gamma}{A}{A}{\idtm{A}}{\idtm{A'}}}
      }
\end{align*}
\pause
Weakening preserves the unit:
\begin{equation*}
\inference
  { \jfam{\Gamma}{A}
    \jfam{{\Gamma}{B}}{Q}
    }
  { \unfoldall{\jhomeq
      {{{\Gamma}{A}}{\ctxwk{A}{B}}}
      {\ctxwk{A}{Q}}
      {\ctxwk{A}{Q}}
      {\ctxwk{A}{\idtm{Q}}}
      {\idtm{\ctxwk{A}{Q}}}
      }
    }
\end{equation*}
\end{frame}
